---
title: "Winter Travel Data Project"
author: "Zack Treisman"
date: "5/27/2021"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(MASS)

by_zone <- read.csv("data/by_zone.csv")
by_zone$rating_below <- factor(by_zone$rating_below)
by_zone$rating_near <- factor(by_zone$rating_near)
by_zone$rating_above <- factor(by_zone$rating_above)
by_zone$modality <- factor(by_zone$modality, levels = c("Non.motorized", "Mechanized", "Hybrid", "Motorized"))
by_zone$Zone <- factor(by_zone$Zone, levels = c("Kebler", "Slate River Rd", "Washington Gulch", "East River", "Brush/Cement"))
```


#  New snow

Most of the patterns in the data relate to user group preferences that don't change much year to year or depending on conditions.

```{r fig.height=5}
ggplot(by_zone, aes(change_depth, user.count, color = Zone))+
  geom_jitter(height = 0, width = 0.2, alpha = 0.5)+
  facet_grid(modality~year)+
  scale_y_log10()+
  labs(x="New Snow (in)", y="# Users")
```

#  Snow in past 3 days

Motorized users head to Kebler, non-motorized users to Snodgrass. Some shift in these patterns year to year is visible.

```{r,fig.height=5}
ggplot(by_zone, aes(past3snow, user.count, color = Zone))+
  geom_jitter(height = 0, width = 0.2, alpha = 0.5)+
  facet_grid(modality~year)+
  scale_y_log10()+
  labs(x="Past 3 Days Snow (in)", y="# Users")
```

#  Air Temperature

Effects of weather and conditions are subtle but not absent. People around here are not too fazed by the cold.

```{r,fig.height=5}
ggplot(by_zone, aes(air_temp, user.count, color = Zone))+
  geom_jitter(height = 0, width = 0.2, alpha = 0.5)+
  facet_grid(modality~year)+
  scale_y_log10()+
  labs(x="Air Temperature (F)", y="# Users")
```


#  CAIC Rating (Near Treeline)

Most activity involving avalanche prone slopes happens near treeline.

```{r, echo=FALSE}
ggplot(na.omit(by_zone), aes(rating_near, user.count, fill=rating_near))+
  geom_jitter(aes(color=rating_near))+
  geom_boxplot(outlier.shape = NA)+
  scale_y_log10()+
  facet_grid(has_sled~Zone)+
  scale_fill_manual(values = c("green", "yellow", "orange", "red"))+
  scale_color_manual(values = c("green", "yellow", "orange", "red"),guide=FALSE)+
  labs(x="", y="# Users", fill="CAIC Rating\n(Near Treeline)")

```


# Effects of weather and danger rating on visitation rates

\scriptsize

```{r}
model1 <- glm.nb(user.count ~ modality*(Zone+rating_near)+air_temp+change_depth+
                   snow_depth+snow_density+weekend, 
                 data=by_zone)
anova(model1)
#summary(model1)
weather_ests <- exp(coef(model1)[c("snow_depth", "change_depth", "snow_density","air_temp")])
rating_ests <- exp(coef(model1)[c("rating_near2", "rating_near3", "rating_near4")])  
highD_ests <- exp(coef(model1)[c("modalityMechanized:rating_near4","modalityHybrid:rating_near4")]) 
m1_CI <- confint(model1)
weather_confs <- exp(m1_CI[c("snow_depth", "change_depth", "snow_density", "air_temp"),])
rating_confs <- exp(m1_CI[c("rating_near2", "rating_near3", "rating_near4"),])
highD_confs <- exp(m1_CI[c("modalityMechanized:rating_near4","modalityHybrid:rating_near4"),])
```

#  Weather and Snow Conditions

Colder temperatures, heavy snow, new snow, and a deep snowpack correlate with decreased visitation. Effects are relatively minor.

```{r, fig.height=5}
weather_model <- data.frame(var.labels=factor(names(weather_ests), levels=names(weather_ests)), weather_ests, 
                        low95 = weather_confs[,1], up95 = weather_confs[,2])
ggplot(weather_model, aes(var.labels, weather_ests))+
  geom_pointrange(aes(ymin=low95, ymax=up95))+
  geom_hline(yintercept=1, linetype = "dashed", color = "red")+
  labs(x = "", y =  "Expected User Count Multiplier per Unit Change")+ coord_flip()
```

#  CAIC Rating (Near Treeline)

Moderate and considerable ratings correlate some with decreased visitation. High danger has more correlation.

```{r, fig.height=5}
rating_model <- data.frame(var.labels=factor(names(rating_ests), levels=names(rating_ests)), rating_ests, 
                        low95 = rating_confs[,1], up95 = rating_confs[,2])
ggplot(rating_model, aes(var.labels, rating_ests))+
  geom_pointrange(aes(ymin=low95, ymax=up95))+
  geom_hline(yintercept=1, linetype = "dashed", color = "red")+
  labs(x = "", y =  "Expected User Count Multiplier")+ coord_flip()
```

#  High ratings and modality

For Mechanized and Hybrid users, the effects of high danger ratings are amplified.

```{r, fig.height=5}
highD_model <- data.frame(var.labels=factor(names(highD_ests), levels=names(highD_ests)), highD_ests, 
                        low95 = highD_confs[,1], up95 = highD_confs[,2])
ggplot(highD_model, aes(var.labels, highD_ests))+
  geom_pointrange(aes(ymin=low95, ymax=up95))+
  geom_hline(yintercept=1, linetype = "dashed", color = "red")+
  labs(x = "", y =  "Expected User Count Multiplier")+ coord_flip()

```


#  Season to season trends

Usage has remained fairly stable, with a definite uptick in non-motorized usage this past season.

```{r, fig.height=5}
ggplot(by_zone, aes(year, user.count))+
  geom_jitter(height = 0, width = 0.2, alpha = 0.7)+
  geom_boxplot()+
  facet_grid(~modality)+
  scale_y_log10()
```

# Season to season trends by user type

Motorized users head to Kebler in increasing numbers, and there were a lot more non-motorized users last year.

```{r, fig.height=5}
year_totals <- by_zone %>%
  group_by(year, Zone, modality) %>%
  summarise(users = sum(user.count, na.rm = TRUE))
ggplot(year_totals, aes(year, users, fill=modality))+
  geom_col()+
  facet_grid(~Zone)+
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  scale_fill_hue(direction=-1)
```

#  Within season trends

There is a jump in usage in the last week of the calendar year, otherwise we see a gradual increase until sometime mid March.

```{r, fig.height=5}
day_totals <- by_zone %>% 
  group_by(date, weekend) %>%
  summarise(users = sum(user.count, na.rm = TRUE)) %>%
  arrange(desc(users))

#head(day_totals, 10)

ggplot(day_totals, aes((yday(date)+30)%%365, users))+
  geom_smooth(span=0.3)+
  geom_point(aes(color=weekend))+
  scale_x_continuous(breaks = c(0,31,62,90,120), 
                     labels = c("Dec 1", "Jan 1", "Feb 1", "Mar 1", "Apr 1"))+
  labs(x="")
```



#  Within season trends by user type

Variance is greater in the non-motorized groups, and the mid March peak is more pronounced for motorized and hybrid users.

```{r, fig.height=5}
day_totals_modes <- by_zone %>% 
  group_by(date, weekend, has_sled) %>%
  summarise(users = sum(user.count, na.rm = TRUE)) %>%
  arrange(desc(users))

ggplot(day_totals_modes, aes((yday(date)+30)%%365, users))+
  geom_point(aes(color=weekend))+
   geom_smooth(span=0.4)+
  scale_x_continuous(breaks = c(0,31,62,90,120), 
                     labels = c("Dec 1", "Jan 1", "Feb 1", "Mar 1", "Apr 1"))+
  #scale_y_log10()+
  labs(x="")+
  facet_grid(~has_sled)
```



